---
title: "Data processing assignment: COVID-19"
output: html_notebook
---


## The data

You are provided with a CSV file, `covid-19.csv` which was downloaded from the EU's open data portal, https://data.europa.eu/euodp/en/home. This file contains data for COVID-19 cases and deaths for each country since the end of 2019.

The dates given in the first column are in the format `dd/mm/yyyy`, e.g. `16/05/2020`. 
These can be converted to R's `date` datatype using the `dmy()` function from the tidyverse `lubridate` package:

```{r}
library(tidyverse)
library(lubridate)

dmy("16/05/2020")
```


## The tasks

0. Use the tidyverse functions to load this file into a dataframe and convert the dates from strings to the `date` datatype. 

```{r}
df <- read_csv('covid-19.csv')
df <- df %>%
  mutate(date=dmy(dateRep))
head(df)
```

---

Now create figures to show the following:

1. Daily cases for a single country.

```{r}
df %>%
  filter(countriesAndTerritories=="Spain") %>%
  ggplot(aes(x=date,y=cases)) +
  geom_line()
```

2. Weekly cases for a single country. *Hint*: use `week()` to convert a date to a number representing the week of the year. `geom_col()` makes a bar chart from (x,y) data.
```{r}

df %>%
  mutate(week=week(date)) %>%
  filter(year==2020 & countriesAndTerritories=="Spain") %>%
  group_by(week) %>%
  summarise(rate = sum(cases)) %>%
  ggplot(aes(x=week,y=rate)) +
  geom_col()
  
```


3. Cumulative daily cases for a single country. *Hint*: use `arrange()` and `cumsum()`


```{r}
df %>%
  filter(countriesAndTerritories=="Spain") %>%
  arrange(date) %>%
  mutate(totalCases = cumsum(cases)) %>%
  ggplot(aes(x=date,y=totalCases)) +
  geom_line() 

```

4. Daily cases for each continent. *Hint*: use `group_by()` and `summarise()`

```{r}
df %>%
  arrange(date) %>%
  group_by(continentExp,date) %>%
  summarise(totalCases = sum(cases)) %>%
  ggplot(aes(x=date,y=totalCases, color=continentExp)) +
  geom_line()
```


5. Apparent overall mortality rate (total deaths / total cases) for each country, grouped by continent.

```{r}

df %>%
  group_by(continentExp,countriesAndTerritories) %>%
  summarise(totalDeaths = sum(deaths), totalCases = sum(cases)) %>%
  mutate(mortality = totalDeaths/totalCases) %>%
  ggplot(aes(y=mortality,x=continentExp, fill=continentExp)) +
  geom_boxplot() + 
  guides(fill='none')

```

---


6. The UK (Referred to as "United_Kingdom" in this data set) has a strong weekly periodicity in reported COVID-19 deaths.

(a) Make a plot to illustrate this effect as clearly as possible. *Hint*: `lubridate` provides a function `wday()` which returns the day of the week as an integer, starting from 1 = Sunday.

```{r}

daymap <- c('other','Monday','Tuesday','other','other','other','other')

df %>%
  filter(countriesAndTerritories=="United_Kingdom") %>%
  filter(date > "2020-03-01") %>%
  mutate(day = wday(date)) %>%
  mutate(day = daymap[day] ) %>%
  ggplot(aes(x=date,y=deaths, fill=day )) +
  geom_col() +                  
  scale_fill_manual(
    values = c('Monday'='red','Tuesday'='blue','other'='grey'),
    breaks = c('Monday','Tuesday','other')  # to order the legend
  )

```

(b) Show that the number of deaths reported on Mondays and Tuesdays is significantly lower than on other days.

```{r}

# We can start by producing a table containing mean daily death rates for each week, grouped into (Monday+Tuesday) vs (other days).

daymap <- c('other','MonTue','MonTue','other','other','other','other')

df2 <- df %>%
  filter(countriesAndTerritories=="United_Kingdom") %>%
  mutate(week=week(date)) %>%
  mutate(day = wday(date)) %>%
  mutate(dayclass = daymap[day] ) %>%
  group_by(week, dayclass) %>%
  summarise(mean_deaths = mean(deaths)) 

df2

```

```{r}

# Let's rearrange this to have the two groups in separate columns. 
# We can also eliminate weeks where either rate is zero.

df3 <- df2 %>%
  spread(key = dayclass, value = mean_deaths) %>%
  filter( MonTue > 0 & other > 0)

df3

```




```{r}

# Owing to the overall shape of the curve, there is a dependency between 
# the mean number of deaths on a (Monday+Tuesday) and the mean number of 
# deaths during the week following. We should therefore use a 
# paired-sample t-test to compare the values in the two columns.

# With significance level alpha=0.05:

t.test(df3$MonTue, df3$other, paired=TRUE, alternative='less' )

```

```{r}

# The resulting p < alpha, so we can reject the null hypothesis that the group 
# (Monday+Tuesday) have the same mean as the other days.

```


(c) Do any other countries show a similar periodicity?


```{r}

# We can use `map_dbl()` to apply the same hypothesis test to all countries:

daymap <- c('other','MonTue','MonTue','other','other','other','other')

do_test <- function(c) {
  
  dfx <- df %>%
    filter(countriesAndTerritories==c) %>%
    mutate(week=week(date)) %>%
    mutate(day = wday(date)) %>%
    mutate(dayclass = daymap[day] ) %>%
    group_by(week, dayclass) %>%
    summarise(mean_deaths = mean(deaths)) %>%
    spread(key = dayclass, value = mean_deaths) %>%
    filter( MonTue > 0 & other > 0)
  
  # Before attempting the t-test, check that
  if( 'MonTue' %in% names(dfx) &       # both groups are present
      'other' %in% names(dfx) &
      nrow(dfx) > 1 &                  # there is more than one week
      ( var(dfx$MonTue) > 0 |          # there is some variation
      var(dfx$other) > 0 )
  ) {
    return( t.test(dfx$MonTue, dfx$other, 
                   paired=TRUE,
                   alternative='less'
                   )$p.value )
  }
  else return( NA )
  
}

all_countries <- df %>%
  filter(deaths > 0) %>%
  pull(countriesAndTerritories) %>%
  sort %>%
  unique

pval <- map_dbl(all_countries, do_test)  # do the tests
names(pval) <- all_countries  # add the country names
pval <- pval[!is.na(pval)]    # remove the countries that couldn't be tested.
pval <- sort(pval)            # sort by ascending p-value

pval[pval< 0.01]             # show the countries with lowest p-value

```


```{r}

# Although we haven't corrected for multiple hypothesis testing, 
# these p-values do appear low enough to be interesting. 

# Let's visualise these distributions:

country_list <- names(pval[pval< 0.01])

daymap <- c('other','Monday','Tuesday','other','other','other','other')

df %>%
  filter(countriesAndTerritories %in% country_list) %>%
  filter(date > "2020-03-01") %>%
  mutate(day = wday(date)) %>%
  mutate(day = daymap[day] ) %>%
  ggplot(aes(x=date,y=deaths, fill=day )) +
  geom_col() +  
  scale_y_log10() +
  facet_wrap(vars(countriesAndTerritories)) +
  scale_fill_manual(
    values = c('Monday'='red','Tuesday'='blue','other'='grey'),
    breaks = c('Monday','Tuesday','other')  # to order the legend
  )




```

```{r}

# We haven't proven anything statistically, but it does seem that other 
# countries experience a similar phenomenon.

```

(d) How could you explain what is happening, and what are the implications for how we use the data?

```{r}

# One possible explanation is that there is an administrative backlog 
# after the weekend (for example, relatives need to be notified before a
# death can be officially reported). This could lead to a reduction in reported 
# deaths early in the week.
#
# The main implication in making use of the data is that it makes no sense to
# compare figures day-by-day, only week-by-week.
#
# You can find more discussion of this sort of periodic effect at
# https://twitter.com/jburnmurdoch/status/1247681939575910402 
# and
# https://jmuccigr.github.io/blog/2020/04/10/Weekly-Periodicity/

```



---
