---
title: "Data processing assignment: Exam examiner"
output: html_notebook
---


## The data

The file `exam_results.xlsx` contains data about some (fictional) students taking modules in a (fictional) Physics department.



## The tasks

0. Load this file into a dataframe. What are the observations and what are the variables?

```{r}
library(tidyverse)
library(readxl)

df <- read_excel('exam_results.xlsx')
df

```

```{r}
#
# At first glance, it looks like we should treat students as observations 
# and each module as a variable, but that would place the grades from both years
# into the same cell (violating rule 3). We might make a decision to aggregate the 
# two years into a single exam result, but that would destroy some potentially 
# important information.
#
# The alternative is to treat each exam result as an observation. This creates a
# lot of redundancy by repeating student data in multiple rows, but that is no problem 
# unless the table were so large that memory became an issue. 
#
# The resulting 9 variables are:
# Student Code, Surname, Forename, Degree (BSc/MSci), Major, Minor, Module, Year, Grade.
#
#
# N.B. 
# The aim of tidying is to hold data in memory in a form from which it is easy to
# generate any tables that we might need for analysis. This is a less stringent 
# kind of normalisation than is needed for database design, where redundancy 
# is avoided at all costs. Making separate tables for Students and Exams would be 
# non-redundant (and still tidy), but leaving them in the same table is fine.
#

```


1. Tidy the data.

```{r}
# Start by gathering the grades into a single column.
# We have to supply the names of the gathered columns as a vector.

df2 <- df %>%
  gather(c('2019','2020'), key='Year', value='Grade', na.rm=TRUE) %>%
  mutate(Year=parse_integer(Year))

df2

```

```{r}
# Next, we need to split the Degree Programme into its component variables.

df3 <- df2 %>%
  separate('Degree Programme', into=c("Degree","Major","with","Minor")) %>%
  select(-with)

df3
```

```{r}
# The data is now tidy!

```

2. Using your tidied data, answer the following questions:

a) For MSci students, what is the mean grade for the Relativity exam?

```{r}
# Use filter to find the right rows, then get the Grade data and take a mean.

df3 %>%
  filter(Degree=='MSci' & Module=='Relativity') %>%
  pull(Grade) %>%
  mean()

```

b) What proportion of the students are taking Physics as their major subject?

```{r}
# We need to make a table with one row per student.

df3 %>%
  select('Student Code','Major') %>%
  unique() %>%
  mutate(Physics=(Major=='Physics')) %>%
  summarise(mean(Physics))

```

```{r}
# Notice that it is much easier to answer these questions using the 
# tidy data than it would have been with the original table!
```



---